---
title: 'Interactions (V)'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    incremental: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, echo = F,message = F, warning = F}
library(tidyverse)
```


## Last time...

Power

---

## Today

- Polynomials

- 3-way interactions

---

## Non-linear relationships

Linear lines often make bad predictions -- very few processes that we study actually have linear relationships. For example, effort had diminishing returns (e.g., log functions), or small advantages early in life can have significant effects on mid-life outcones (e.g., exponentional functions). In cases where the direction of the effect is constant but changing in magnitude, the best way to handle the data is to transform a variable (usually the outcome) and run linear analyses.

```{r, eval = F}
log_y = ln(y)
lm(log_y ~ x)
```
---

## Polynomial relationships

Other processes represent changes in the directon of relationship -- for example, it is believed that a small amount of anxiety is benefitial for performance on some tasks (positive direction) but too much is detrimental (negative direction). When the shape of the effect includes change(s) in direction, then a polynomial term(s) may be more appropriate.

It should be noted that polynomials are often a poor approximation for a non-linear effect of X on Y. However, correctly testing for non-linear effects usually requires (a) a lot of data and (b) making a number of assumptions about the data. Polynomial regression can be a useful tool for exploratory analysis and in cases when data are limited in terms of quantity and/or quality.


---
## Polynomial regression 
 
Polynomial regression is most often a form of hierarchical regressoin that systematically tests a series of higher order functions for a single variable.

$$
\begin{aligned}
\large \textbf{Linear: } \hat{Y} &= b_0 + b_1X \\
\large \textbf{Quadtratic: } \hat{Y} &= b_0 + b_1X + b_2X^2 \\
\large \textbf{Cubic: } \hat{Y} &= b_0 + b_1X + b_2X^2 + b_3X^3\\
\end{aligned}
$$

---
### Example

Can a team have too much talent? Researchers hypothesized that teams with too many talented players have poor intrateam coordination and therefore perform worse than teams with a moderate amount of talent. To test this hypothesis, they looked at 208 international football teams. Talent was the percentage of players during the 2010 and 2014 World Cup Qualifications phases who also had contracts with elite club teams. Performance was the number of points the team earned during these same qualification phases.

```{r, message=F}
library(here)
football = read.csv(here("data/swaab.csv"))
```

.small[Swaab, R.I., Schaerer, M, Anicich, E.M., Ronay, R., and Galinsky, A.D. (2014). [The too-much-talent effect: Team
interdependence determines when more talent is too much or not enough.](https://www8.gsb.columbia.edu/cbs-directory/sites/cbs-directory/files/publications/Too%20much%20talent%20PS.pdf) _Psychological Science 25_(8), 1581-1591.]

---
.pull-left[
```{r, message = F, warning = F}
head(football)
```
]
.pull-right[
```{r, warning = F}
ggplot(football, aes(x = talent, y = points)) +
  geom_point() + 
  geom_smooth(se = F) +
  theme_bw(base_size = 20)
```

]


---

```{r mod1, warning = F, message = F, fig.width = 10, fig.height = 5}
mod1 = lm(points ~ talent, data = football)
library(broom)
aug1 = augment(mod1)
ggplot(aug1, aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_smooth(se = F) +
  theme_bw(base_size = 20)
```

---
```{r mod2}
mod2 = lm(points ~ talent + I(talent^2), data = football)
summary(mod2)
```

---

```{r, message = F}
library(sjPlot)
plot_model(mod2, type = "pred")
```

---
## Interpretation

The intercept is the predicted value of Y when x = 0 -- this is always the interpretation of the intercept, no matter what kind of regression model you're running. 

The $b_1$ coefficient is the tangent to the curve when X=0. In other words, this is the rate of change when X is equal to 0. If 0 is not a meaningful value on your X, you may want to center, as this will tell you the rate of change at the mean of X.


```{r}
football$talent_c50 = football$talent - 50
mod2_50 = lm(points ~ talent_c50 + I(talent_c50^2), data = football)
```

---

```{r}
summary(mod2_50)
```

---
```{r, echo = FALSE, warning= FALSE}
plot_model(mod2_50, type = "pred")
```
---

Or you can choose another value to center your predictor on, if there's a value that has a particular meaning or interpretation.

```{r}
football$talent_c = football$talent - mean(football$talent)
mod2_c = lm(points ~ talent_c + I(talent_c^2), data = football)
```

---

```{r}
mean(football$talent)
summary(mod2_c)
```

---
```{r, echo = FALSE, warning= FALSE}
plot_model(mod2_c, type = "pred")
```
---


## Interpretation

The $b_2$ coefficient indexes the acceleration, which is how much the slope is going to change. More specifically, $2 \times b_2$ is the acceleration: the rate of change in $b_1$ for a 1-unit change in X.

You can use this to calculate the slope of the tangent line at any value of X you're interested in:
$$\large b_1 + (2\times b_2\times X)$$

---

```{r}
tidy(mod2)
```

.pull-left[
**At X = 10**
```{r}
54.9 + (2*-.570*10)
```

]
.pull-right[
**At X = 70**
```{r}
54.9 + (2*-.570*70)
```

]

---

## Polynomials are interactions

An term for $X^2$ is a term for $X \times X$ or the multiplication of two independent variables holding the same values. 

```{r}
football$talent_2 = football$talent*football$talent
tidy(lm(points ~ talent + talent_2, data = football))
```
---

## Polynomials are interactions

Put another way:

$$\large \hat{Y} = b_0 + b_1X + b_2X^2$$

$$\large \hat{Y} = b_0 + \frac{b_1}{2}X + \frac{b_1}{2}X + b_2(X \times X)$$

The interaction term in another model would be interpreted as "how does the slope of X change as I move up in Z?" -- here, we ask "how does the slope of X change as we move up in X?"
---

## When should you use polynomial terms?

You may choose to fit a polynomial term after looking at a scatterplot of the data or looking at residual plots. A U-shaped curve may be indicative that you need to fit a quadratic form -- although, as we discussed before, you may actually be measuring a different kind of non-linear relationship. 

Polynomial terms should mostly be dictated by theory -- if you don't have a good reason for thinking there will be a change in sign, then a polynomial is not right for you.

And, of course, if you fit a polynomial regression, be sure to once again check your diagnostics before interpreting the coefficients. 
---

```{r, warning = F, message = F, fig.width=10, fig.height=5}
aug2 = augment(mod2)
ggplot(aug2, aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_smooth(se = F) +
  theme_bw(base_size = 20)

```

---

```{r, warning = F, message = F, fig.width=10, fig.height=7}
plot_model(mod2, type = "pred", show.data = T)
```

---

class: inverse

## Three-way interactions and beyond

---


### Three-way interactions (regression)

**Regression equation**

$$\hat{Y} = b_{0} + b_{1}X + b_{2}Z + b_{3}W + b_{4}XZ + b_{5}XW + b_{6}ZW + b_{7}XZW$$
The three-way interaction qualifies the three main effects (and any two-way interactions). 

Like a two-way interaction, the three-way interaction is a conditional effect.  And it is symmetrical, meaning there are several equally correct ways of interpreting it.

**Factorial ANOVA**

We describe the factorial ANOVA design by the number of levels of each factor. "X by Z by W" (e.g., 2 by 3 by 4, or 2x3x4)
---
A two-way  (A x B) interaction means that the magnitude of one main effect (e.g., A main effect) depends on levels of the other variable (B).  But, it is equally correct to say that the magnitude of the B main effect depends on levels of A. In regression, we refer to these as **conditional effects** and in ANOVA, they are called **simple main effects.**

A three-way interaction means that the magnitude of one two-way interaction (e.g., A x B) depends on levels of the remaining variable (C). 

It is equally correct to say that the magnitude of the A x C interaction depend on levels of B. Or, that the magnitude of the B x C interaction depends on levels of A. These are known as **simple interaction effects**.

---
### Example (regression)

```{r, echo = F}
set.seed(03032020)
gender = sample(c("male", "female"), size = 150, replace = T)
bad_day = sample(x = c(1:5), size = 150, replace = T)
talk = round(2*(gender == "female") + .2*(bad_day) + rnorm(n = 150))
talk = talk + 1
talk[talk < 1] = 1
talk[talk > 5] = 5
stress = bad_day + talk - (bad_day*talk) + round(rnorm(n = 150, sd = 5))
stress_m = bad_day + talk + (bad_day*talk)+ round(rnorm(n = 150, sd = 5))
stress[gender == "male"] = stress_m[gender == "male"]
stress = stress+1+abs(min(stress))

stress_data = data.frame(gender, bad_day, talk, stress)
```


```{r}
psych::describe(stress_data, fast = T)
table(stress_data$gender)
```

---

```{r}
mod_stress = lm(stress ~ bad_day*talk*gender, data = stress_data)
summary(mod_stress)
```

---

```{r, message = F}
library(reghelper)
simple_slopes(mod_stress)
```

---

```{r, fig.width=10, fig.height = 7}
plot_model(mod_stress, type = "int", mdrt.values = "meansd")
```

---

As a reminder, centering will change all but the highest-order terms in a model. 

```{r}
stress_data = stress_data %>%
  mutate(bad_day_c = bad_day - mean(bad_day),
         talk_c = talk - mean(talk))
newmod = lm(stress ~ bad_day_c*talk_c*gender, data = stress_data)
```

---
```{r}
tidy(mod_stress)
```
```{r}
tidy(newmod)
```


---
### Four-way? 


$$
\begin{aligned}
\hat{Y} &= b_0 + b_1X + b_{2}Z + b_{3}W + b_{4}Q + b_{5}XW\\
&+ b_{6}ZW + b_{7}XZ + b_{8}QX + b_{9}QZ + b_{10}QW\\
&+ b_{11}XZQ + b_{12}XZW + b_{13}XWQ + b_{14}ZWQ + b_{15}XZWQ\\
\end{aligned}
$$

--

3-way (and higher) interactions are incredibly difficult to interpret, in part because they represent incredibly complicated processes. If you have a solid theoretical rationale for conducting a 3-day interaction, be sure you've collected enough subjects to power your test (more on this later). 

---

Especially with small samples, three-way interactions may be the result of a few outliers skewing a regression line. If you have stumbled upon a three-way interactoin during exploratory analyses, **be careful.** This is far more likely to be a result of over-fitting than uncovering a true underlying process.

Best practice for 3-way (and higher) interactions is to use at least one nominal moderator (ideally with only 2 levels), instead of all continuous moderators. This allows you to examine the 2-way interaction at each level of the nominal moderator.

Even better if one of these moderators is experimenter manipulated, which increases the likelihood of balanced conditions. 

---

class: inverse

## Next time...

Statistical power and wrapping up interactions