---
title: 'Bootstrapping'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    incremental: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

#Last time...

Wrapping up interactions

* Power

# Today

Robust estimation
* Weighted Least Squares
* Boostrapping

---

ADD WLS Here
* violation of homoscedasticity

---

WLS is a robust method for dealing with heteroscedasticity. What if you violate the normality assumption of regression?

--

A **bootstrapping** approach is useful when the theoretical sampling distribution for an estimate is unknown or unverifiable. In other words, if you have reason to suspect that either the variables in your model are not normally distributed, or that they're non-normally distributed, then bootstrapping can help ensure non-bias estimation and appropriately sized confidence intervals. 

---

## Bootstrapping

In bootstrapping, the theoretical sampling distribution is assumed to be unknown or unverifiable. Under the weak assumption that the sample in hand is representative of some population, then that population sampling distribution can be built empirically by randomly sampling with replacement from the sample.

The resulting empirical sampling distribution can be used to construct confidence intervals and make inferences.
---
### Example

A sample of 216 response times. What is their central tendency and variability?

There are several candidates for central tendency (e.g., mean, median) and for variability (e.g., standard deviation, interquartile range).  Some of these do not have well understood theoretical sampling distributions.

For the mean and standard deviation, we have theoretical sampling distributions to help us, provided we think the mean and standard deviation are the best indices. For the others, we can use bootstrapping.

---
```{r, echo = F, message=F, fig.width=10, fig.height=8}
library(tidyverse)
set.seed(03102020)
response = rf(n = 216, 3, 50) 
response = response*500 + rnorm(n = 216, mean = 200, sd = 100)

values = quantile(response, 
                  probs = c(.025, .5, .975))
mean_res = mean(response)

data.frame(x = response) %>%
  ggplot(aes(x)) +
  geom_histogram(aes(y = ..density..), binwidth = 150, fill = "lightgrey", color = 
                   "black")+
  geom_density()+
  geom_vline(aes(xintercept = values[1], 
                 color = "Lower 2.5%"), size = 2)+
  geom_vline(aes(xintercept = values[2], 
                 color = "Median"), size = 2)+
  geom_vline(aes(xintercept = values[3], 
                 color = "Upper 2.5%"), size = 2)+
  geom_vline(aes(xintercept = mean_res, 
                 color = "Mean"), size = 2)+
  labs(x = "Reponse time (ms)", title = "Response Time Distribution") + cowplot::theme_cowplot(font_size = 20)
```

---
### Bootstrapping

Before now, if we wanted to estimate the mean and the 95% confidence interval around the mean, we would find the theoretical sampling distribution by scaling a t-distribution to be centered on the mean of our sample and have a standard deviation equal to $\frac{s}{\sqrt{N}}.$ But we have to make many assumptions to use this sampling distribution, and we may have good reason not to.  

Instead, we can build a population sampling distribution empirically by randomly sampling with replacement from the sample.

---

### Illustration

Imagine you had a sample of 6 people: Rachel, Monica, Phoebe, Joey, Chandler, and Ross. To bootstrap their heights, you would draw from this group many samples of 6 people *with replacement*, each time calculating the average height of the sample.

```{r, echo = F}
friends = c("Rachel", "Monica", "Phoebe", "Joey", "Chandler", "Ross")
heights = c(65, 65, 68, 70, 72, 73)
names(heights) = friends
(sample1 = sample(friends, size = 6, replace = T)); mean(heights[sample1])
(sample1 = sample(friends, size = 6, replace = T)); mean(heights[sample1])
(sample1 = sample(friends, size = 6, replace = T)); mean(heights[sample1])
(sample1 = sample(friends, size = 6, replace = T)); mean(heights[sample1])
(sample1 = sample(friends, size = 6, replace = T)); mean(heights[sample1])
(sample1 = sample(friends, size = 6, replace = T)); mean(heights[sample1])
(sample1 = sample(friends, size = 6, replace = T)); mean(heights[sample1])
(sample1 = sample(friends, size = 6, replace = T)); mean(heights[sample1])
```

---

```{r}
boot = 10000
response_means = numeric(length = boot)
for(i in 1:boot){
  sample_response = sample(response, size = 216, replace = T)
  response_means[i] = mean(sample_response)
}
```

```{r, echo = F, message = F, fig.width = 10, fig.height=6}
data.frame(means = response_means) %>%
  ggplot(aes(x = means)) + 
  geom_histogram(color = "white") +
  geom_density() +
  geom_vline(aes(xintercept = mean(response_means), color = "mean"), size = 2) +
  geom_vline(aes(xintercept = median(response_means), color = "median"), size = 2) +
  geom_vline(aes(xintercept = quantile(response_means, probs = .025), color = "Lower 2.5%"), size = 2) +
    geom_vline(aes(xintercept = quantile(response_means, probs = .975), color = "Upper 2.5%"), size = 2) +
  cowplot::theme_cowplot()
```


---

```{r}
mean(response_means)
median(response_means)

quantile(response_means, probs = c(.025, .975))
```

What about something like the median?

---

```{r}
boot = 10000
response_med = numeric(length = boot)
for(i in 1:boot){
  sample_response = sample(response, size = 216, replace = T)
  response_med[i] = median(sample_response)
}
```
.pull-left[
```{r echo=F,  message=FALSE}
data.frame(medians = response_med) %>%
  ggplot(aes(x = medians)) + 
  geom_histogram(aes(y = ..density..),
                 color = "white", fill = "grey") +
  geom_density() +
  geom_vline(aes(xintercept = mean(response_med), color = "mean"), size = 2) +
  geom_vline(aes(xintercept = median(response_med), color = "median"), size = 2) +
  geom_vline(aes(xintercept = quantile(response_med, probs = .025), color = "Lower 2.5%"), size = 2) +
    geom_vline(aes(xintercept = quantile(response_med, probs = .975), color = "Upper 2.5%"), size = 2) +
  cowplot::theme_cowplot()
```
]
.pull-right[
```{r}
mean(response_med)
median(response_med)
quantile(response_med, 
         probs = c(.025, .975))
```
]
---

```{r}
boot = 10000
response_sd = numeric(length = boot)
for(i in 1:boot){
  sample_response = sample(response, size = 216, replace = T)
  response_sd[i] = sd(sample_response)
}
```
.pull-left[
```{r echo=F,  message=FALSE}
data.frame(sds = response_sd) %>%
  ggplot(aes(x = sds)) + 
  geom_histogram(aes(y = ..density..),color = "white", fill = "grey") +
  geom_density() +
  geom_vline(aes(xintercept = mean(response_sd), color = "mean"), size = 2) +
  geom_vline(aes(xintercept = median(response_sd), color = "median"), size = 2) +
  geom_vline(aes(xintercept = quantile(response_sd, probs = .025), color = "Lower 2.5%"), size = 2) +
    geom_vline(aes(xintercept = quantile(response_sd, probs = .975), color = "Upper 2.5%"), size = 2) +
  cowplot::theme_cowplot()
```
]
.pull-right[
```{r}
mean(response_sd)
median(response_sd)
quantile(response_sd, 
         probs = c(.025, .975))
```
]

---

You can bootstrap estimates and 95% confidence intervals for *any* statistics you'll need to estimate. 

The `boot` function provides some functions to speed this process along.

```{r}
library(boot)

# function to obtain R-Squared from the data
rsq <- function(formula, data, indices) {
  d <- data[indices,] # allows boot to select sample
  fit <- lm(formula, data=d)
  return(summary(fit)$r.square)
}
# bootstrapping with 10000 replications
results <- boot(data=mtcars, statistic=rsq,
   R=10000, formula=mpg~wt+disp)
```

---
.pull-left[
```{r}
data.frame(rsq = results$t) %>%
  ggplot(aes(x = rsq)) +
  geom_histogram(color = "white", bins = 30) 
```
]

.pull-right[
```{r}
median(results$t)
boot.ci(results, type = "perc")
```
]

---

### Example 2

Samples of service waiting times for Verizon’s (ILEC) versus other carriers (CLEC) customers. In this district, Verizon must provide line service to all customers or else face a fine. The question is whether the non-Verizon customers are getting ignored or facing greater variability in waiting times.

```{r, message = F, warning = F, echo = 2}
library(here)
Verizon = read.csv(here("data/Verizon.csv"))
```

```{r, echo = F, fig.width = 10, fig.height = 4}
Verizon %>%
  ggplot(aes(x = Time, fill = Group)) + 
  geom_histogram(bins = 30) + 
  guides(fill = F) +
  facet_wrap(~Group, scales = "free_y")
```

---

```{r, echo = F, fig.width = 10, fig.height = 6}
Verizon %>%
  ggplot(aes(x = Time, fill = Group)) + 
  geom_histogram(bins = 50, position = "dodge") + 
  guides(fill = F)
table(Verizon$Group)
```

---

There's no world in which these data meet the typical assumptions of an independent samples t-test. To estimate mean differences we can use boostrapping. Here, we'll resample with replacement separately from the two samples. 

```{r}
boot = 10000
difference = numeric(length = boot)

subsample_CLEC = Verizon %>% filter(Group == "CLEC")
subsample_ILEC = Verizon %>% filter(Group == "ILEC")

for(i in 1:boot){
  sample_CLEC = sample(subsample_CLEC$Time, size = nrow(subsample_CLEC), replace = T)
  sample_ILEC = sample(subsample_ILEC$Time, size = nrow(subsample_ILEC), replace = T)
  
  difference[i] = mean(sample_CLEC) - mean(sample_ILEC)
}
```

---

```{r echo=F,  message=FALSE, fig.width=10, fig.height=5}
data.frame(differences = difference) %>%
  ggplot(aes(x = differences)) + 
  geom_histogram(aes(y = ..density..),color = "white", fill = "grey") +
  geom_density() +
  geom_vline(aes(xintercept = mean(differences), color = "mean"), size = 2) +
  geom_vline(aes(xintercept = median(differences), color = "median"), size = 2) +
  geom_vline(aes(xintercept = quantile(differences, probs = .025), color = "Lower 2.5%"), size = 2) +
    geom_vline(aes(xintercept = quantile(differences, probs = .975), color = "Upper 2.5%"), size = 2) +
  cowplot::theme_cowplot()
```

The difference in means is `r round(median(difference),2)` $[`r round(quantile(difference, probs = .025),2)`,`r round(quantile(difference, probs = .975),2)`]$.

---

### Bootstrapping Summary

Bootstrapping can be a useful tool to estimate parameters when 
1. you've violated assumptions of the test (i.e., normality, homoskedasticity)
2. you have good reason to believe the sampling distribution is not normal, but don't know what it is
3. there are other oddities in your data, like very unbalanced samples 

This allows you to create a confidence interval around any statistic you want -- Cronbach's alpha, ICC, Mahalanobis Distance, $R^2$, AUC, etc. 
* You can test whether these statistics are significantly different from any other value -- how?

---

### Bootstrapping Summary

Bootstrapping will NOT help you deal with:

* dependence between observations -- for this, you'll need to explicity model dependence (e.g., multilevel model, repeated measures ANOVA)

* improperly specified models or forms -- use theory to guide you here

* measurement error -- why bother?

---

## Yarkoni and Westfall (2017)

Y&W describe the goals of explanation and prediction in science; how are these goals similar to each other and how are they in opposition to each other? 

According to Y&W, how should psychologists change their research, in terms of explanation and prediction, and why?

How do regression models fit into the goals of explanation and prediction? Where do they fall short on one or other or both?

???

Explanation: describe causal underpinnings of behaviors/outcomes
Prediction: accurately forecast behaviors/outcomes

Similar: both goals of science; good prediction can help us develop theory of explanation and vice versa

Statistical tension with one another: statistical models that accurately describe causal truths often have poor prediction and are complex; predictive models are often very different from the data-generating processes. 

Y&W: we should spend more time and resources developing predictive models than we do not (not necessarily than explanation models, although they probably think that's true)

---

## Yarkoni and Westfall (2017)

What is **overfitting** and where does this occur in terms of the models we have discussed in class thus far?

What is **bias** and **variance**, and how does the bias-variance trade-off relate to overfitting?

* How concerned are you about overfitting in your own area of research? How about in the studies you'd like to do in the next couple of years?

???

Overfitting: mistakenly fitting sample-specific noise as if it were signal

OLS models tend to be overfit because they minimize error for a specific sample

Bias: systematically over or under estimating parameters
Variance: how much estimates tend to jump around

Model-fits tend to prioritizie minimizing bias or variance, and choosing to minimize one inflates the other; OLS models minimize one of these


---

## Yarkoni and Westfall (2017)

How do Y&W propose adjusting our current statistical practices to be more successful at prediction? 

???

big data sets

Cross-validation

regularization

---

## Yarkoni and Westfall (2017)


**Big Data**
* Reduce the likelihood of overfitting -- more data means less error

**Cross-validation**
* Is my model overfit?

**Regularization**
* Constrain the model to be less overfit 

---

### Big Data Sets

"Every pattern that could be observed in a given dataset reflects some... unknown combination of signal and error" (page 1104). 

Error is random, so it cannot correlate with anything; as we aggregate many pieces of information together, we reduce error. 

Thus, as we get bigger and bigger datasets, the amount of error we have gets smaller and smaller

---

### Cross-validation

**Cross-validation** is a family of techniques that involve testing and training a model on different samples of data. 
* Replication
* Hold-out samples
* K-fold
    * Split the original dataset into 2(+) datasets, train a model on one set, test it in the other
    * Recycle: each dataset can be a training AND a testing; average model fit results to get better estimate of fit
    * Can split the dataset into more than 2 sections
    
---

```{r, message=FALSE, warning = F}
library(here)
stress.data = read.csv(here("data/stress.csv"))
library(psych)
describe(stress.data, fast = T)

model.lm = lm(Stress ~ Anxiety*Support*group, 
              data = stress.data)
summary(model.lm)$r.squared
```

---

### Example: 10-fold cross validation

```{r, message = F}
# new package!
library(caret)
# set control parameters
ctrl <- trainControl(method="cv", number=10)
# use train() instead of lm()
cv.model <- train(Stress ~ Anxiety*Support*group, 
               data = stress.data, 
               trControl=ctrl, # what are the control parameters
               method="lm") # what kind of model
cv.model
```

---

### Regularization

Penalizing a model as it grows more complex. 
* Usually involves shrinking coefficient estimates -- the model will fit less well in-sample but may be more predictive

*lasso regression*: balance minimizing sum of squared residuals (OLS) and minimizing smallest sum of absolute values of coefficients
- coefficients are more biased (tend to underestimate coefficients) but produce less variability in results
