---
title: 'Univariate regression II'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    incremental: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Last time...

- Introduction to univariate regression

- Calculation and interpretation of $b_0$ and $b_1$

- Relationship between $X$, $Y$, $\hat{Y}$, and $e$ 

---

### Today...

Statistical inferences with regression

- Partitioning the variance

- Testing $b_{xy}$

---

## Statistical Inference

- The way the world is = our model + error

- How good is our model? Does it "fit" the data well? 

--

To assess how well our model fits the data, we simply take all the variability in our outcome and partition it into different categories. For now, we will partition it into two categories: the variability that is predicted by (explained by) our model, and variability that is not.

---

## Partitioning variance (variability)

When it comes to partitioning variance, we often use the sum of the squared deviations for the math.

For one, minimizing the total residual squared deviations is the goal of OLS. 

For another, squared deviations are additive, making the math relatively simple. The total squared deviations of Y is equal to the total residual squared deviations plus the variance of $\hat{Y}$:

$$\Large \sum (Y - \bar{Y})^2 = \sum (\hat{Y} -\bar{Y})^2 + \sum(Y - \hat{Y})^2$$

---

$$\Large \sum (Y - \bar{Y})^2 = \sum (\hat{Y} -\bar{Y})^2 + \sum(Y - \hat{Y})^2$$

For simplicity and readability, we may abbreviate these:

$$\Large SS_Y = SS_{\text{Model}} + SS_{\text{Residual}}$$

The relative magnitude of sums of squares, especially in more complex designs, provides a way of identifying particularly large and important sources of variability. In the future, we can further partition $SS_{\text{Model}}$ and $SS_{\text{Residual}}$ into smaller pieces, which will help us make more specific inferences and increase statistical power, respectively. 

$$\Large s^2_Y = s^2_{\hat{Y}} + s^2_{e}$$
---

## Partitioning variance in Y
- Consider the case with no correlation between X and Y

$$\Large \hat{Y} = \bar{Y} + r_{xy} \frac{s_{y}}{s_{x}}(X-\bar{X})$$
$$\Large \hat{Y} = \bar{Y}$$

- To the extent that we can generate different predicted values of Y based on the values of the predictors, we are doing well in our prediction

$$\Large \sum (Y - \bar{Y})^2 = \sum (\hat{Y} -\bar{Y})^2 + \sum(Y - \hat{Y})^2$$

$$\Large SS_Y = SS_{\text{Model}} + SS_{\text{Residual}}$$

---
## Coefficient of Determination

$$\Large \frac{s_{Model}^2}{s_{y}^2} = \frac{SS_{Model}}{SS_{Y}} = R^2$$


---
## Example
```{r, echo = 2:4}
galton.data = psychTools::galton
fit.1 <- lm(child ~ parent, data = galton.data)
summary(fit.1)
summary(fit.1)$r.squared
```

---
## Example

```{r}
cor.test(galton.data$parent, galton.data$child)
```

--

```{r}
cor(galton.data$parent, galton.data$child)^2
```

---
## Computing Sum of Squares

$$\Large \frac{SS_{Model}}{SS_{Y}} = R^2$$
$$\Large SS_{Model} = R^2({SS_{Y})}$$
$$\Large SS_{residual} = SS_{Y} - R^2({SS_{Y})}$$

$$\Large  SS_{residual} = (1- R^2){SS_{Y}}$$
???
 
```{r,echo=FALSE, message=FALSE, warning = F}
set.seed(123)
library(broom)
library(tidyverse)
x.1 <- rnorm(10, 0, 1)
e.1 <- rnorm(10, 0, 2)
y.1 <- .5 + .55 * x.1 + e.1
d.1 <- data.frame(x.1,y.1)
m.1 <- lm(y.1 ~ x.1, data = d.1)
d1.f<- augment(m.1)
d1.f$color = "black"
d1.f[9, "color"] = "red"
newcolors = c("black", "red")
```


```{r, echo=FALSE}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
  geom_point(size = 2, aes(color = color)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point(aes(y = .fitted, color = color), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted, color = color))+
  geom_hline(aes(yintercept = mean(y.1)))+
  scale_color_manual(values = newcolors)+
  guides(color = F)+
  theme_bw(base_size = 20)
```

---
## residual standard error

```{r, echo=FALSE}
summary(fit.1)
```

---
## residual standard error/deviation 

- aka standard deviation of the residual
- aka standard error of the estimate

$$\Large \hat{\sigma_e}$$

- interpreted in original units (unlike $R^2$)
- standard deviation of Y not accounted by model

---

## residual standard error/deviation  


```{r}
summary(fit.1)$sigma 
galton.data.1 = broom::augment(fit.1)
psych::describe(galton.data.1$.resid)
```

---
## residual standard error/deviation 

```{r}
summary(fit.1)$sigma 
psych::describe(galton.data$parent)
```
---


```{r, echo = F}
set.seed(20200116)
x.1 = rnorm(1000,0,1)
e = seq(4, .01, by = -.1)
r2 = numeric(length = length(e))
rsd= numeric(length = length(e))
for(i in 1:length(e)){
  e.1 <- rnorm(1000, 0, e[i])
  y.1 <- .5 + .55 * x.1 + e.1
  d.1 <- data.frame(x.1,y.1)
  m.1 <- lm(y.1 ~ x.1, data = d.1)
  r2[i] = summary(m.1)$r.squared
  rsd[i] = summary(m.1)$sigma
}

data.frame(r2 = r2, rsd = rsd) %>%
  ggplot(aes(x = r2, y = rsd)) + 
  geom_line() +
  scale_x_continuous(expression(paste(R^2))) + 
  scale_y_continuous(expression(hat(sigma_e))) 
```
---
## $R^2$ and residual standard deviation
- two sides of same coin
- one in original units, the other standardized 
- $R^2$ can be tricky because the numerator and denominator can be changed in different ways. 
- for example if variance in Y is changed but with the same model and residual standard error $R^2$ could decline or increase

---
## Inferential tests

NHST is about making decisions:
  
  - these two means are/are not different
  - this correlation is/is not significant
  - the distribution of this categorical variabe is/is not different between these groups
  
--

In regression, there are several inferential tests being conducted at once. The first is called the **omnibus test** -- this is a test of whether the model fits the data. 

---

### Omnibus test

$$\Large H_{0}: \rho_{XY}^2= 0$$

$$\Large H_{0}: \rho_{XY}^2 \neq 0$$

It is possible to calculate the significance of your regression with a correlation test. In fact, it would seem quite practical and logical to do so. 

--

However, historically we use a different probability distribution -- **the _F_ distribution** to estimate the signifiance of our model. It's important to know that these methods are mathematically equivalent. But the _F_ distribution is useful here, because it works with our ability to partition variance.

---

```{r}
data.frame(x = c(0, 5)) %>%
  ggplot(aes(x = x)) +
  stat_function(fun = function(x) df(x, df1 = 3, df2 = 10), 
                geom = "line", aes(color = "df1 = 3", linetype = "df2 = 10")) +
  stat_function(fun = function(x) df(x, df1 = 1, df2 = 10), 
                geom = "line", aes(color = "df1 = 1", linetype = "df2 = 10")) +
  stat_function(fun = function(x) df(x, df1 = 5, df2 = 10), 
                geom = "line", aes(color = "df1 = 5", linetype = "df2 = 10")) +
    stat_function(fun = function(x) df(x, df1 = 3, df2 = 50), 
                geom = "line", aes(color = "df1 = 3", linetype = "df2 = 50")) +
  stat_function(fun = function(x) df(x, df1 = 1, df2 = 50), 
                geom = "line", aes(color = "df1 = 1", linetype = "df2 = 50")) +
  stat_function(fun = function(x) df(x, df1 = 5, df2 = 50), 
                geom = "line", aes(color = "df1 = 5", linetype = "df2 = 50")) +
  scale_y_continuous("density")+
  theme_bw()
```


---

## _F_ Distribution review

The F probability distribution represents all possible ratios of two variances:

$$F \approx \frac{s^2_{1}}{s^2_{2}}$$
Each variance estimate in the ratio is $\chi^2$ distributed, if the data are normally distributed. The ratio of two $\chi^2$ distributed variables is $F$ distributed. It should be noted that each $\chi^2$ distribution has its own degrees of freedom.

$$F_{\nu_1\nu_2} = \frac{\frac{\chi^2_{\nu_1}}{\nu_1}}{\frac{\chi^2_{\nu_2}}{\nu_2}}$$
As a result, _F_ has two degrees of freedom, $\nu_1$ and $\nu_2$
---

$$\Large F = \frac{MS_{Model}}{MS_{residial}}$$
---
## model comparison

- Last semester you looked at different models to see how their F and SSs changed depending on what predictors were included. We can do the same thing with regression models!   

- The basic idea is asking how much variance remains unexplained in our model. This "left over" variance can be contrasted with an alternative model/hypothesis. We can ask does adding a new predictor variable help explain more variance or should we stick wtih a parsimonious model.   

- Every model test you do implicitly implies you favoring that over an alternative model, typically the null. This framework allows you to be more flexible and explicit.   

---
```{r}
fit.1 <- lm(parent ~ child, data = galton.data)
fit.0 <- lm(parent ~ 1, data = galton.data)
```


---
```{r}
summary(fit.0)
```

---
```{r}
summary(fit.1)
```


---
```{r}
anova(fit.0)
```

---
```{r}
anova(fit.1)
```


---
```{r}
anova(fit.1, fit.0)
```
---
## model comparisons 

- Model comparisons are redundent with nil/null hypotheses and coefficient tests right now, but will be more flexible down the road. 
- Key is to start thikning about your implicit alternative models
- The ultimate goal would be to create two models that represent two equally plausible theories. 
- Theory A is made up of components XYZ, whereas theory B has QRS components. You can then ask which theory(model) is better? 
---
## regression coefficient

$$\Large H_{0}: \beta_{1}= 0$$
$$\Large H_{1}: \beta_{1} \neq 0$$
---
## What does the regression coefficient test?
- Does X provide any predictive information? 
- Does X provide any explanatory power regarding the variability of Y? 
- Is the the average value the best guess (i.e., is Y bar equal to the predicted value of Y?)
- Is the regression line flat? 
- Are X and Y correlated?  
---
## Regression coefficient
$$\Large se_{b} = \frac{s_{Y}}{s_{X}}{\sqrt{\frac {1-r_{xy}^2}{n-2}}}$$
$$\Large t(n-2) = \frac{b_{1}}{se_{b}}$$
** what is standardized equation? 
---
## Intercept

- same idea, more complex se calculation as the calculation depends on how far the X value (here zero) is away from the mean of X
- farther from the mean, less information, thus more uncertainty 
- we will come back and see this equation later
---
## Confidence interval for coefficents

- same equation as we've been working with
- estimate plus minus 1.96*se
---
## Confidence Bands for regression line
```{r, echo=FALSE, message=FALSE}
set.seed(123)

px.1 <- rnorm(1000, 0, 1)
pe.1 <- rnorm(1000, 0, 1)
py.1 <- .5 + .55 * px.1 + pe.1
pd.1 <- data.frame(px.1,py.1)

px.2 <- rnorm(100, 0, 1)
pe.2 <- rnorm(100, 0, 1)
py.2 <- .5 + .55 * px.2 + pe.2
pd.2 <- data.frame(px.2,py.2)


p1 <- ggplot(pd.1, aes(x = px.1,y =  py.1)) +
    geom_point() +    
    geom_smooth(method = lm) +
   scale_x_continuous(limits = c(-3, 3))  +
  scale_y_continuous(limits = c(-3, 3))

p2 <- ggplot(pd.2, aes(x=px.2, y=py.2)) +
    geom_point() +    
    geom_smooth(method=lm) +
  scale_x_continuous(limits = c(-3, 3))  +
  scale_y_continuous(limits = c(-3, 3))

library(cowplot)
plot_grid(p1, p2, ncol=2, labels = c("N = 1000", "N = 100"))

```
---
## Confidence Bands
- Compare mean estimate for height of 70 based on regression vs binning
- Model uses all data where binning uses much less

```{r, echo=FALSE}
ggplot(galton.data, aes(x=child, y=parent)) +
      geom_point() +   
  geom_smooth(method=lm,   # Add linear regression line
                se=TRUE) 
```
---
## Confidence Bands

$$\Large \hat{Y}\pm t_{critical} * se_{residual}*\sqrt{\frac {1}{n}+\frac{(X-\bar{X})^2}{(n-1)s_{X}^2}} $$


---
## Prediction
- very similar to confidence bands around regressions line
- differences is that we are predicting and individual i's score, not the Y hat for a particular level of X. (A new Yi given x, rather than Ybar given x)
- Because there is greater variation in predicting an individual value rather than a collection of individual values (ie the mean) the prediction band is greater
- Combines unknown variability in 1) the estimated mean (as reflected in se of b) 2) peoples scores around  mean (residual standard error) 

$$\Large \hat{Y}\pm t_{critical} * se_{residual}*\sqrt{1+ \frac {1}{n}+\frac{(X-\bar{X})^2}{(n-1)s_{X}^2}} $$

---

```{r, warning=FALSE}
temp_var <- predict(fit.1, interval="prediction")
new_df <- cbind(galton.data, temp_var)
pred <- ggplot(new_df, aes(x=child, y=parent))+
       geom_point() +   
  geom_smooth(method=lm,se=TRUE) +
 geom_ribbon(aes(ymin = lwr, ymax = upr), 
               fill = "blue", alpha = 0.1)
```


---
```{r}
pred
```

---
## Mean square error (MSE)


- AKA means square residual/within

- unbiased estimate of error variance

- measure of discrepancy between the data and the model

- the MSE is the variance around the fitted regression line

- just like $\text{MS}_{\text{within}}$ was variance around predicted group means

---
## standard errors for b
- represent our uncertainty (noise) in our estimate of the regression coefficient 
- different from residual standard error/deviation (but proportional to)  
- much like previously we can take our estimate (b) and put confidence regions around it to get an estimate of what could be "possible" if we ran the study again  
- (see equation later)
