---
title: "Lab 9: Bootstrapping"
output: 
  html_document: 
    fig_caption: yes
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
    df_print: paged
editor_options: 
  chunk_output_type: console
---

You will need the Rmd file for today's lab. You can download it [here](https://uopsych.github.io/psy612/labs/lab-9/lab-9.Rmd).

```{r setup, include=FALSE}
# suppress scientific notation
options(scipen = 999)

# do not evaluate code chunks
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE)
```

## Libraries

Be sure to have the following packages installed and loaded:

```{r lab-8-1, message=FALSE}
library(tidyverse) # for plotting and data wrangling
library(rio)       # for importing data
library(boot)      # for bootstrapping
library(psych)
```

## Data

We'll work with that Lab 2 dataset we worked with before. Recall that it has:

1. extraversion: self-reported extraversion
2. pos_express: self-reported positive expressivity
3. soc_sup: self-reported social support
4. optimism: self-reported optimism
5. exercise: self-reported exercise
6. happiness: self-reported happiness

```{r}
df <- import("https://raw.githubusercontent.com/uopsych/psy612/master/labs/lab-2/data/PSY612_Lab2_Data.csv") %>%
  janitor::clean_names()
```


And we'll also work with the a dataframe we used in the *regression diagnostics* lab (`df_het`) that demonstrated an issue with heteroskedasticity. It has only the variables `extraversion` and `happiness`.
```{r}
df_het <- import(file     = "https://raw.githubusercontent.com/uopsych/psy612/master/labs/lab-6/data/df_het.csv",
              setclass = "tibble") %>%
  janitor::clean_names()
```


# Purpose

Bootstrapping is a procedure for empirically determining a sampling distribution when we cannot assume a statistic follows a known distribution. Specifically, we resample with replacement from our own sample to determine how a statistic is distributed when using the current data.

This lab covers how to use bootstrapping in R using the `{boot}` library. 

1. [Why bootstrap?](#why)
2. [Bootstrapping with `{boot}`](#boot)
3. [Writing a function for `boot()`](#functionsBoot)
4. [Running `boot()`](#runBoot)
5. [Another bootstrapping example](#example)



# Why Bootstrap?{#why}

Bootstrapping is a general purpose tool to use whenever you can't rely on assumptions about how a statistic is distributed. Some cases where it is useful include:   

1. When test assumptions  (e.g., normality, homoscedasticity) have been violated  
2. When you don't know what the shape of a statistic's sampling distribution (or have good reason to believe it is non-normal)

3. With highly unbalanced sample sizes 

It is often used to construct CIs based on bootstrapped samples rather than CIs that rely on assumptions about the data.


For example, recall from the regression diagnostics lab that we ran into a model that demonstrated heteroskedasticity.
```{r}
model <- lm(happiness ~ extraversion, data = df_het)
plot(model, 1) # get residuals plot 
```

Ordinary least squares calculates standard errors based on the assumption that they are approximately equally distributed across the range of our model (homoscedasticity assumption).

```{r}
summary(model)
```

Thus, the standard errors calculated for our model coefficients are calculated based on an assumption that we can see we have violated. Instead of using SEs that rely on a model assumption we've violated, we can estimate SEs from the data using the **bootstrapping** method.


# Bootstrapping with `boot()` {#boot}

We're going to focus on using the `boot()` function from the `{boot}` package. `boot()` has many arguments, but there are three that are necessary:   
   
1. `data` = This is where you tell `boot()` the name of the dataframe you're working with.   

2. `statistic` = a __function__ that returns the desired statistic when applied to the data.

3. `R` = the number of resamples. You usually want to set it to a very large number (e.g., 10,000).

And that's it. The only part that is a little tricky is specifying argument 2, because you have to *write your own function* to get the statistic you're looking for. This may be new to all of you, but writing functions in R is a powerful skill and so it is something that is definitely worth practicing. 


# Writing a function for `boot()` {#functionsBoot}

For `boot()`, we want to write a function that returns the statistic we're interested in bootstrapping. Basically, you want to set up the body such that:

It has two necessary arguments that are (in order) `data` and `index`. The `data` argument comes first, so whatever you name it will represent the dataframe in the body of the function. The `index` argument comes second, so whatever you name it should be used to subset the data. 

Let's write a function that runs our model (`lm(happiness ~ extraversion)`) and extracts from the model output the parameter estimate corresponding to extraversion (`b_ext`).
```{r}
# your code here
get_b_ext <- function(data, index){
  
  data_subset <- data[index,]
  
  model <- lm(happiness ~ extraversion, data = data_subset)
  
  b_ext <- summary(model)$coefficients[2,1]
  
  return(b_ext)
}
```

# Running `boot()` {#runBoot}

Now, we have the function that `boot()` can use to bootstrap this statistic we chose. Let's go ahead and give it a try. We set (in order) the `data`, `statistic`, and `R` arguments

- The data argument is where we provide the name of the dataset.
- The statistic argument is where we provide the name of the function we created.
- R is the number of times that we want to sample with replacement

```{r}
set.seed(101) # don't forget to set.seed()

boot_b_ext <- boot(data   = df_het, 
                   statistic = get_b_ext,
                   R         = 10000)
```


## Getting Useful Information from `boot()` Results

The most important object in the list returned by `boot()` is `t`, which is a matrix that contains our bootstrapped results. It will be the length of `R`, so in this case it will be length 10,000 (i.e., 10,000 rows); this means there are 10,000 estimates of $b_ext$, each pertaining to the estimate obtained in one of the resamplings. 

Let's look at just the first ten estimates of b_ext:
```{r}
boot_b_ext$t[1:10,]
```

Let's also see what the mean and SD of the b_ext values are:
```{r}
mean(boot_b_ext$t)
sd(boot_b_ext$t)
```

And let's look at what the resulting object contains:
```{r}
boot_b_ext
```

The `original` value refers to the estimate you would have obtained if you had not used bootstrapping, the `bias` value refers to the difference between the bootstrapped estimate and the estimate you would have obtained if you had not used bootstrapping, and `std.error` refers to the standard deviation of the sample distribution.


## Plotting Bootstrapped Results

We can also generate plots to describe our bootstrapped distribution. Let's do that using `{ggplot2}`. The only sort of tricky part is that we have to turn the bootstrapped results into a dataframe or tibble first. I've done this below using the `tibble()` function from the `{tidyverse}`

First, let's create a density plot

```{r}
# create a density 
tibble(boot_b_ext = boot_b_ext$t) %>%                   
  ggplot(aes(x = boot_b_ext)) +                      
    geom_density(fill = "turquoise", alpha = .6) + 
    theme_bw()                                     
```

And, we can get the 95% CI by using `boot.ci()`, passing the results of `boot()` as the first argument, and setting `type = "perc"` for percentiles.

```{r}
ci <- boot.ci(boot_b_ext, type = "perc")
ci
```



# Another bootstrapping example {#example}

The above was an example of bootstrapping a statistic when assumptions about the model have been violated. Let's also go over an example of bootstrapping a statistic to estimate its standard error when the shape of its sampling distribution is unknown or non-normally distributed, like the sampling distribution of $R^2$.

First, let's examine the other data set.
```{r}
head(df)
str(df)
```

Let's use these variables to construct a model predicting happiness from all other variables in the data frame.
```{r}
model <- lm(happiness ~ extraversion + pos_express + soc_sup + optimism + exercise, data = df)
summary(model)
```

What if we wanted to know the standard error associated with the $R^2$ value? Notice there is none given in the output. Instead, we can estimate the standard error by bootstrapping this statistic.

First, write a function called `get_r_sq` that will run the above model and extract the R^2 value from the output:
```{r}
# your code here
get_r_sq <- function(data, index){
  
  data_subsetted <- data[index,] 
  
  fit            <- lm(happiness ~ extraversion + 
                                   pos_express  + 
                                   soc_sup      + 
                                   optimism     + 
                                   exercise, 
                      data = data_subsetted)
  
  rsq <- summary(fit)$r.squared
  
  return(rsq)
}
```

Next, use that function in the `boot` function to resample from the data set 10,000 times, obtaining an estimate of $R^2$ each time:
```{r}
set.seed(21) # don't forget to set.seed()

boot_r_sq <- boot(data      = df,
                  statistic = get_r_sq,
                  R         = 10000)
```

Examine the mean and SD of the $R^2$ values you obtained through bootstrapping:
```{r}
# your code here
mean(boot_r_sq$t)
sd(boot_r_sq$t)
```

Plot the distribution of $R^2$ values:
```{r}
# your code here
tibble(boot_r_sq = boot_r_sq$t) %>%                   
  ggplot(aes(x = boot_r_sq)) +                      
    geom_density(fill = "turquoise", alpha = .6) + 
    theme_bw()
```


Obtain a 95%CI around the mean estimate of the $R^2$ value:
```{r}
# your code here
ci <- boot.ci(boot_r_sq, type = "perc")
ci
```


# More advanced applications {#advanced}

You can also add additional arguments that you want to provide as inputs to your statistic-grabbing function. 

For example, say we want to specify a different model each time:
```{r}
get_r_sq <- function(data, index, formula){
  
  data_subset <- data[index,] 
  
  model <- lm(formula, data = data_subset)
  
  rsq <- summary(fit)$r.squared
  
  return(rsq)
}
```

Then, when we use our boot function, we can run a number of different models to construct a sampling distribution of $R^2$ values from:
```{r}
boot_r2_model1 <- boot(data      = df,     
                       statistic = get_r_sq, 
                       formula   = happiness ~ extraversion,
                       R         = 1000) 

boot_r2_model2 <- boot(data      = df,     
                       statistic = get_r_sq, 
                       formula   = happiness ~ extraversion + pos_express,
                       R         = 1000) 

boot_r2_model3 <- boot(data      = df,     
                       statistic = get_r_sq, 
                       formula   = happiness ~ extraversion + pos_express + soc_sup,
                       R         = 1000) 

boot_r2_model4 <- boot(data      = df,     
                       statistic = get_r_sq, 
                       formula   = happiness ~ extraversion + pos_express + soc_sup + optimism,
                       R         = 1000) 

boot_r2_model5 <- boot(data      = df,     
                       statistic = get_r_sq, 
                       formula   = happiness ~ extraversion + pos_express + soc_sup + optimism + exercise,
                       R         = 1000)
```


Let's see some cool visualiaztions we can do with these five different sampling distributions. First, let's organize the output of each one (the sampled $R^2$ values) and the names of the predictors for each model into separate tables:
```{r}
r2_model1_results <- tibble(rsq        = boot_r2_model1$t,
                            predictors = "Extr.")

r2_model2_results <- tibble(rsq        = boot_r2_model2$t,
                            predictors = "Extr. + Pos. Expr.")

r2_model3_results <- tibble(rsq        = boot_r2_model3$t,
                            predictors = "Extr. + Pos. Expr. + Soc. Supp.")

r2_model4_results <- tibble(rsq        = boot_r2_model4$t,
                            predictors = "Extr. + Pos. Expr. + Soc. Supp. + Opt.")

r2_model5_results <- tibble(rsq        = boot_r2_model5$t,
                            predictors = "Extr. + Pos. Expr. + Soc. Supp. + Opt. + Exercise")
```

Combine these results for each model into a single data frame. Also, calculate the medians and CIs for each of the models:

```{r}
r2_all_models <- r2_model1_results %>%
  rbind(r2_model2_results) %>% 
  rbind(r2_model3_results) %>% 
  rbind(r2_model4_results) %>% 
  rbind(r2_model5_results)
```


We could show each of the sampling distributions side-by-side using `geom_density`, and fill the color of the density geoms according to the model they correspond to. Let's take a look at what that looks like:

```{r}
ggplot(r2_all_models, aes(x = rsq, fill = predictors)) + 
  geom_density(alpha = .5) +
  labs(x = expression(paste("Bootstrapped estimates of ", R^{2}))) +
  guides(fill = guide_legend(nrow = 3)) +
  scale_fill_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")
```



# Minihacks

## Minihack 1

Using the same dataset above, calculate bootstrapped estimates of *adjusted* $R^2$ value:   

```{r}
# your code here
get_r2 <- function(data, index, formula){
  
  data_subsetted <- data[index,] 
  
  fit            <- lm(formula,
                      data = data_subsetted)
  
  rsq <- summary(fit)$adj.r.squared
  
  return(rsq)
}

set.seed(42)

boot_adjr2_model5 <- boot(data      = df,     
                          statistic = get_r2, 
                          formula   = happiness ~ extraversion + pos_express + soc_sup + optimism + exercise,
                          R         = 1000) 
```

## Minihack 2

Get bootstrapped estimates of correlations between:

1. Extraversion and Positive Expressivity   
2. Extraversion and Social Support   
3. Extraversion and Optimism   
4. Extraversion and Exercise   

Then, construct a plot of the distribution for each one.

```{r}
# your code here
get_r <- function(data, index, vars){
  
  data_subsetted <- data[index,]
  
  r              <- cor(data_subsetted[,vars])[2, 1]
  
  return(r) 
}

set.seed(42)

cor_extr_posexp <- boot(data      = df,  
                        statistic = get_r, 
                        vars      = c("extraversion", "pos_express"),
                        R         = 1000)

cor_extr_socsup <- boot(data      = df,  
                        statistic = get_r, 
                        vars      = c("extraversion", "soc_sup"),
                        R         = 1000) 

cor_extr_opt    <- boot(data      = df,  
                        statistic = get_r, 
                        vars      = c("extraversion", "optimism"),
                        R         = 1000)

cor_extr_exer <- boot(data      = df,  
                      statistic = get_r, 
                      vars      = c("extraversion", "exercise"), 
                      R         = 1000)


cor_extr_posexp_results <- tibble(r    = cor_extr_posexp$t,
                                  vars = "Extr., Pos. Expr.")

cor_extr_socsup_results <- tibble(r    = cor_extr_socsup$t,
                                  vars = "Extr., Soc. Supp.")

cor_extr_opt_results    <- tibble(r    = cor_extr_opt$t,
                                  vars = "Extr., Opt.")

cor_extr_exer_results   <- tibble(r    = cor_extr_exer$t,
                                  vars = "Extr., Exercise")

cor_all <- cor_extr_posexp_results %>% 
  rbind(cor_extr_socsup_results) %>% 
  rbind(cor_extr_opt_results) %>% 
  rbind(cor_extr_exer_results) %>% 
  group_by(vars) %>% 
  summarize(median = median(r),
            lower  = quantile(r, probs = .025),
            upper  = quantile(r, probs = .975))

ggplot(cor_all, aes(x = vars, y = r)) +
  geom_pointrange(aes(y = median, ymin = lower, ymax = upper),
                  color = "turquoise3") +
  coord_flip() +
  labs(x = NULL,
       y = expression(paste("Zero-order ", italic(r), " (with bootstrapped 95% CIs)"))) +
  theme_bw()
```