---
title: "Lab 8: Factorial ANOVA"
output: 
  html_document: 
    fig_caption: yes
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
    df_print: paged
---

```{r setup, include=FALSE}
# suppress scientific notation
options(scipen = 999)
```


# Purpose
XXX


1. [Research scenario](#scenario)



Be sure to have the following packages installed and loaded:

```{r message=FALSE}
library(tidyverse) # for plotting and data wrangling
library(rio) # for importing data
library(psych) # for descriptives
```

***

# Research scenario{#scenario}

Based on subjects' self-reports of rejection sensitivity (N = 80), a researcher divides subjects into two equal groups (low RS and high RS).  Whereas half of the subjects in each group interact with a partner who displays a happy emotional expression during the interaction, the other half of the subjects in each group interact with a partner who displays a neutral emotional expression during the interaction.  After the interaction, subjects are asked to rate the statement, my interaction partner likes me, on a scale from 1 (strongly disagree) to 7 (strongly agree).

## Data

```{r}
reject <- import("https://raw.githubusercontent.com/uopsych/psy612/master/labs/lab-8/data/reject.csv")
```

